Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-28 paper/paper.tex
write abstract and clean up bibliography

\todo{
\begin{abstract}
    Write abstract.
\end{abstract}
}

@@

for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting. A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e. $\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}. However, in our setting we often cannot guarantee either of the two conditions. Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.

>>>>

\begin{abstract}
    Stochastic trace estimation is a family of well established techniques for approximating the trace of constant matrices. However, oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$. This is the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$. However, this approach does not scale well if $m$ is large. We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization for all values of the parameter $t$. We show that all of these estimators closely match their corresponding error guarantees for constant matrices. In particular, we show that for the Nyström++ method, we can guarantee an error of order $\varepsilon$ by just using $\mathcal{O}(\varepsilon^{-1})$ random vectors to build the estimate in all values of $t$, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has. We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method. Finally, we confirm our theoretical developments in multiple numerical experiments.
\end{abstract}

@@

for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a symmetric parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting. A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e. $\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}. However, in our setting we often cannot guarantee either of the two conditions. Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.
